(Re-)bridging the anatomy ontologies with SSSOM

### 2024-06-14 presentation.tex
Move some slides to the appendix.

  But to enable \emph{cross-species} curation and analysis of data (especially
  high-throughput data, such as scRNAseq), we need ideally a single,
  multi-species ontology that covers all our model organisms. For the past 12
  years or so, this has been provided by Uberon, the Uber Anatomy Ontology.

  Rather than duplicating the work already done by pre-existing,
  species-specific ontologies, Uberon's strategy is to provide a taxon-neutral
  core framework, describing common anatomical structures in broadly generic
  terms, and to plug the species-specific ontologies into that core.}

@@

  This is the role of cross-ontology \emph{mappings} -- basically, a mapping
  is a statement that one term from an ontology A is ``mapped'', or
  ``related'', to another term in an ontology B.}

@@

  Furthermore, the custom script that generates the bridge files from the
  cross-references is highly specific to Uberon; its behaviour are completely
  hardcoded to fit Uberon's use case, so the integration logic cannot possibly
  be reused by other projects.}

@@

  First, we want to treat mappings as ``first-class entities'', on the same
  level as the terms of the ontology. Among other things, this means that they
  must be publicly available on their own, without requiring people interested
  in them to forcefully extract them from the ontology.

  Second, we want to store the mappings using a format that is both standard
  and specifically intended for representing mappings. That format should,
  among other things, make it possible to store mapping-specific metadata, and
  to give precise meaning to a mapping (not just saying that two terms are
  mapped, but what is the nature of the link between them).

@@

  \note{One of the benefits of the SSSOM standard is that it does not simply
  allow to say that two terms are mapped, or somehow related, to each other.
  As we've seen in the last example, it also allows to specify an explicit
  mapping predicate, so as to give a precise meaning to each mapping. It is
  then possible to distinguish, for example between mappings of terms that
  represent exactly the same concept in two different vocabularies, and
  mappings of terms that represent \emph{similar}, but not exactly identical,
  concepts.

  In fact, as part of the efforts around the SSSOM standard, we have compiled
  a dedicated vocabulary, the \emph{Semantic Mapping Vocabulary}, to host,
  among other things, various mapping predicates that can be used to describe
  mappings. In that vocabulary, as part of this project, we have added a
  handful of predicates to specifically represent \emph{cross-species
  mappings}.

  For example, \emph{crossSpeciesExactMatch}, which is intended for mappings
  between terms that represent the same concept, modulo a taxonomic
  restriction. We can use that predicate to explicitly say that the FBbt term
  for the fly heart represents exactly the same concept as the Uberon term for
  dorsal vessel heart, \emph{except} for the fact that the former is specific
  to flies.}

@@

  \note<1>{Now that we can store mappings in a dedicated format, we need a way
  to use those mappings when we want to integrate several ontologies together.
  And, as stated in our goals, we want to do that in a way that is, as much as
  possible, not specific to Uberon, but reusable by other projects as well.

@@

@@

  \note{This leads us to our new, SSSOM-based integration pipeline for Uberon,
  which now has three distinct inputs: (1) Uberon itself, (2) the mappings
  between Uberon and the taxon-specific ontologies, and (3) the taxon-specific
  ontologies.

  \note{For Uberon itself, the main benefit of the new approach is that we now
  have a pipeline that is both more flexible --~for example, it allows
  combining mapping sets from different sources as input~-- and more robust.

>>>>

  But to enable \emph{cross-species} curation and analysis of data (especially
  high-throughput data, such as scRNAseq), we need ideally a single,
  multi-species ontology that covers all our model organisms. For the past 12
  years or so, this has been the goal of Uberon, the Uber Anatomy Ontology.

  Uberon's strategy to achieve that goal is to provide a taxon-neutral core
  framework, describing common anatomical structures in broadly generic terms,
  and to plug the species-specific ontologies into that core.}

@@

  This is the role of cross-ontology \emph{mappings}.}

@@

  Furthermore, the custom script that generates the bridge files from the
  cross-references is highly specific to Uberon; its behaviour are completely
  hardcoded to fit Uberon's use case, so the integration logic is difficult to
  modify and almost impossible to reuse by other projects.}

@@

  First, we want to treat mappings as ``first-class entities'', on the same
  level as the terms of the ontology. This notably means that they must be
  publicly available on their own, without requiring people interested in them
  to forcefully extract them from the ontology.

  Second, we want to store the mappings using a format that is both standard
  and specifically intended for representing mappings. That format should make
  it possible to store mapping-specific metadata, and to give precise meaning
  to a mapping (not just saying that two terms are mapped, but: who asserted
  that those two terms should be mapped? When? On what basis? And what does
  the mapping actually means?).

@@

  \note{One of the benefits of the SSSOM standard is that it does not simply
  allow to say that two terms are mapped, to each other.  As we've seen in the
  last example, it also allows to specify an explicit mapping predicate, so as
  to give a precise meaning to each mapping. It is then possible to
  distinguish, for example between mappings of terms that represent exactly
  the same concept in two different vocabularies, and mappings of terms that
  represent \emph{similar}, but not exactly identical, concepts.

  The authors of the SSSOM standard have compiled a dedicated vocabulary, the
  \emph{Semantic Mapping Vocabulary}, to host, among other things, various
  mapping predicates that can be used to describe mappings. In that
  vocabulary, as part of this project, we have added a handful of predicates
  to specifically represent \emph{cross-species mappings}.

  The most important being \emph{crossSpeciesExactMatch}, which is intended
  for mappings between terms that represent the same concept, modulo a
  taxonomic restriction. We can use that predicate to explicitly say that the
  FBbt term for the fly heart represents exactly the same concept as the
  Uberon term for dorsal vessel heart, \emph{except} for the fact that the
  former is specific to flies.}

@@

  \note<1>{Now that we can store mappings in a dedicated format, we need a way
  to use those mappings when we want to integrate several ontologies together.
  And, as per our third goal, we want to do that in a way that is, as much as
  possible, not specific to Uberon, but reusable by other projects as well.

@@

  \note{This leads us to our new, SSSOM-based integration pipeline for Uberon.
  That pipeline has three distinct inputs: (1) Uberon itself, (2) the mappings
  between Uberon and the taxon-specific ontologies (which are no longer
  embedded within the ontology itself), and (3) the taxon-specific ontologies.

  That step is performed by a dedicated ROBOT plugins that implement the
  SSSOM/T-OWL language.

  \note{For Uberon itself, the main benefit of the new approach is that we now
  have a pipeline that is both more flexible --~for example, easy to modify,
  as all the logic is in the SSSOM/T-OWL ruleset~-- and more robust.

\appendix
\section{\appendixname}

@@
